DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2022-05-10 manuscript/manuscript.Rmd
addressing more comments and adding citations

Here we present the DateLife project and its core software application, available as an R package [@datelifeR], and as an online Rshiny interactive website at www.datelife.org, which captures data from published chronograms, and make these data readily accessible to users for reuse and reanalysis.
The software features key elements for scientific reproducibility, such as a versioned, open and fully public chronogram database [@mctavish2015phylesystem], age data stored in a computer readable format [@vos2012nexml], automated and programmatic ways of accessing the data [@Stoltzfus2013] and methods to summarize and compare age data.

@@

DateLife's core software application consists of the R package `datelife`. Its latest stable version -- v`r packageVersion("datelife")`, is available from the CRAN repository [@datelifeR], and relies on functionalities from various biological R packages:
ape [@paradis2004],
bold [@Chamberlain2018],
geiger [@Harmon2008],
paleotree [@Bapst2012a],
phyloch [@Heibl2008],
phylocomr [@Ooms2018],
phytools [@Revell2012],
rotl [@Michonneau2016], and
taxize [@Chamberlain2013; @Chamberlain2018].
Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: providing an input, searching a chronogram database, and summarizing results from the search.

<!--_**Providing an input.--**_ -->

## Processing an input

DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree. If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004]. The input tree is not required to have branch lengths, and its topology is used in the summary steps described below.

@@

## Searching the database

@@

We illustrate the DateLife algorithm using a group within the Passeriform birds encompassing the family of true finches, Fringillidae as case study. The first example analyses 6 bird species and shows all steps of the algorithm. The second example is an application analysing all 289 species in the family Fringillidae for which we have date informtaion.
<!--We used `r cite_r("r-references.bib")` for all our analyses.-->

@@

We randomly chose 6 bird species in the family Fringillidae, the true finches. The sample includes
two species of cardinals:
the black-thighed grosbeak -- *Pheucticus tibialis* and
the crimson-collared grosbeak -- *Rhodothraupis celaeno*;
three species of buntings:
the yellowhammer -- *Emberiza citrinella*,
the pine bunting --  *Emberiza leucocephalos* and
the yellow-throated bunting --  *Emberiza elegans*;
and one species of tanager, the vegetarian finch -- *Platyspiza crassirostris*.

Processing input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the default reference taxonomy [Open Tree of Life Taxonomy v3.3, June 1, 2021]. For a detailed discussion on the state of the synonym refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza]. 
DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies. Three studies matched five input names [@barker2015new; @Hedges2015; @Jetz2012], one study matched four input names [@hooper2017chromosomal] and two studies matched two input names [@barker2013going; @burns2014phylogenetics]. No studies matched all input names. Together, matching chronograms have 28 unique age data points. All nodes have age data.
DateLife used OpenTree's synthetic tree topology for these taxa and mapped age data to nodes in the tree.
As expected, more inclusive nodes (e.g., node "n1") have more age data than less inclusive nodes (e.g., node "n5").
The processing step allowed discovering five data points for node "n4" that would not have had any data otherwise. \textcolor{red}{I don't understand what this sentence means}
Age summary statistics per node were calculated and used as secondary calibrations to date the tree topology using the BLADJ algorithm.
Age data for node "n2" was excluded as final calibration because it is older than age data of a more inclusive node (n1).

@@

To estimate ages for the taxa in the family Fringillidae, we ran a `datelife` query using the "get species from taxon" flag. 
This will get all recognised species names within the named taxon. 
The Fringillidae has species, according to the Open Tree of Life taxonomy.
??? We have date estimates for `r length(dquery$cleaned_names)` of those species.

??? There is an option to include all, right?
According to the OpenTree of life synthetic phylogeny, Fringillidae is not a monophyletic family, so this taxon-constrained approach returns a topology and ages for the species in that named group.

The next step is to run a `datelife` search on this curated set of species taxon names that will find all chronograms that contain at least two species. The algorithm proceeds to prune the trees to keep matching species names on tips only, and transform the pruned trees to pairwise distance matrices.
There are `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
The final step is to summarize the available information using two alternative types of summary chronograms, median and SDM. As explained in the "Description" section, data from source chronograms is first summarised into a single distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig. \ref{fig:fringillidages}). Median sxummary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
<!-- ????????????????? Say some things about the results!
ANSWER:
(Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
<!-- and hierarchical clustering method which appears to push ages back -->

@@

We found that node ages from original study, and ages estimated using all other age data available are correlated  \textcolor{red}{R^2 value?} (Supplementary Fig. \ref{fig:cvXY}).

In 5 studies, more inclusive nodes have older original ages, and less inclusive nodes have younger original ages than their cross-validated age estimates. 
\textcolor{red}{more inclusive nodes have to be older! :P I think there is something differnet you are trying to say here - maybe "Datelife tended to underestimate ages for more inclusive nodes relative to the orginal estimate, and overestimate ages for nodes closer to the tips"}
Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. \ref{fig:cvXYdiffs}).

>>>>

Here we present the DateLife project and its core software application, available as an R package [@datelifeR], and as an online Rshiny interactive website at www.datelife.org, which captures data from published chronograms, and make these data readily accessible to users for reuse and reanalysis.
The software features key elements for scientific reproducibility, such as a versioned, open and fully public source database [@mctavish2015phylesystem], data stored and available in a computer readable format [@vos2012nexml], automated and programmatic ways of accessing the data [@Stoltzfus2013] and methods to summarize and compare the data.

@@

DateLife's core software application consists of the R package `datelife`. Its latest stable version -- v`r packageVersion("datelife")`, is available from the CRAN repository [@datelifeR], and relies on functionalities from various biological R packages:
ape [@paradis2004],
bold [@Chamberlain2018],
geiger [@Harmon2008],
paleotree [@Bapst2012a],
phyloch [@Heibl2008],
phylocomr [@Ooms2018],
phytools [@Revell2012],
rotl [@Michonneau2016], and
taxize [@Chamberlain2013; @Chamberlain2018].
Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: creating a search query, searching a database, and summarizing results from the search.

<!--_**Processing an input.--**_ -->

##  Creating a search query

DateLife starts by processing a user input consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree. If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004]. The input tree is not required to have branch lengths, and its topology is used in the summary steps described below.

@@

## Searching a chronogram database

@@

We illustrate the DateLife algorithm using a group within the Passeriform birds known as the family of true finches, Fringillidae, as case study. The first example analyses 6 bird species and shows all steps of the algorithm. The second example is an application analysing 289 species in the family Fringillidae included in the NCBI taxonomy.
<!--We used `r cite_r("r-references.bib")` for all our analyses.-->

@@

_**Creating a search query.-**_
We chose 6 bird species in the family Fringillidae, known as the true finches. The sample includes
two species of cardinals:
the black-thighed grosbeak -- *Pheucticus tibialis* and
the crimson-collared grosbeak -- *Rhodothraupis celaeno*;
three species of buntings:
the yellowhammer -- *Emberiza citrinella*,
the pine bunting --  *Emberiza leucocephalos* and
the yellow-throated bunting --  *Emberiza elegans*;
and one species of tanager, the vegetarian finch -- *Platyspiza crassirostris*.

Processing input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the default reference taxonomy [Open Tree of Life Taxonomy v3.3, June 1, 2021]. For a detailed discussion on the state of the synonym refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza]. 
Discovering this synonym allowed assigning five age data points for the parent node of *Emberiza elegans*, shown as *Schoeniclus elegans* in figure \ref{fig:figure2-1}, which would not have had any data otherwise. 
<!--\textcolor{red}{I don't understand what this sentence means}
it is related to the synonym found by tnrs processing, if we had used the synonym insted of the one accepted in the database, we would not have discovered that information
It moved it here, it makes more sense here.
-->

_**Searching the database.-**_
DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies. Three studies matched five input names [@barker2015new; @Hedges2015; @Jetz2012], one study matched four input names [@hooper2017chromosomal] and two studies matched two input names [@barker2013going; @burns2014phylogenetics]. No studies matched all input names. Together, matching chronograms have 28 unique age data points. All nodes have age data.

_**Summarizing search results.-**_
DateLife used OpenTree's synthetic tree topology for these taxa and mapped age data to nodes in the tree.
As expected, more inclusive nodes (e.g., node "n1") have more age data than less inclusive nodes (e.g., node "n5").
Age summary statistics per node were calculated and used as secondary calibrations to date the tree topology using the BLADJ algorithm.
Age data for node "n2" was excluded as final calibration because it is older than age data of the more inclusive node "n1".

@@

_**Creating a query.-**_
To estimate ages for species in the family Fringillidae, we ran a `datelife` query using the "get species from taxon" flag, 
which gets all recognized species names within a named group from a taxonomy of choice (options are Open Tree of Life, NCBI, GBIF, or IRMNG). 
Following the NCBI taxonomy, the Fringillidae has `r length(dquery$cleaned_names)` species.

_**Searching the database.-**_
The next step is to use this curated set of species taxon names to identify all chronograms that contain at least two Fringillidae species. Once identified, the algorithm proceeds to prune matching chronograms to keep Fringillidae species names on tips only, and transform these pruned chronograms to pairwise distance matrices.
The `datelife` search revealed `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].

_**Summarizing search results.-**_
The final step is to summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM. As explained in the "Description" section, a tree topology to summarize age data upon is required.
By default, `datelife` uses the Open Tree of Life synthetic phylogeny as summarizing topology. According to this phylogeny, Fringillidae is not a monophyletic family 
[@burns2014phylogenetics;
@burleigh2015building;
@Jetz2012;
@lerner2011multilocus;
@lovette2010comprehensive;
@pulgarin2013multilocus;
@chaves2013biogeography;
@lamichhaney2015evolution;
@parchman2007coevolution;
@selvatti2015paleogene;
@moyle2016tectonic;
@odeen2011evolution;
@claramunt2015new;
@beresford2005african;
@kimball2019phylogenomic;
@barker2015new;
@klicka2014comprehensive;
@barker2013going;
@treplin2008molecular;
@price2014niche;
@hackett2008phylogenomic;
@barker2014mitogenomic;
@gibb2015new;
@barker2004phylogeny;
@zuccon2012phylogenetic;
@oliveros2019earth;
@johansson2008phylogenetic;
@alstrom2014discovery;
@powell2014comprehensive;
@bryson2014diversification;
@packert2012horizontal;
@tietze2013complete].

Hence, DateLife's taxon-constrained approach returns a topology and ages for the species in a named group.
Age data from source chronograms is summarised into a single distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig. \ref{fig:fringillidages}).

Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
<!-- ????????????????? Say some things about the results!
ANSWER:
(Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
<!-- and hierarchical clustering method which appears to push ages back -->

@@

We found that node ages from original study, and ages estimated using all other age data available are correlated (Supplementary Fig. \ref{fig:cvXY}).

In 5 studies, DateLife presented a tendency to underestimate ages for more inclusive nodes, and overestimate ages for nodes closer to the tips, relative to the original age estimates.
<!-- i.e., more inclusive nodes have older original ages, and less inclusive nodes have younger original ages than their cross-validated age estimates. 
\textcolor{red}{more inclusive nodes have to be older!}
-->
Accordingly, root ages estimated with DateLife's cross-validated ages are generally younger relative to the age root from the original study  (Supplementary Fig. \ref{fig:cvXYdiffs}).
